---
title: "oceanmap Package Tutorial"
subtitle: 'Visualizing and Analyzing Spatial Oceanographic Data'

author:
- name: "Dr. Robert Bauer | Data Scientist & Fishery Biologist"
  affiliation: 1

address:
- code: 1
  address: Hawai'i Institute of Marine Biology, University of Hawai'i at Mānoa, Kāne'ohe, Hawai'i, United States of America
  email: rkbauer@hawaii.edu
  orcid: https://orcid.org/0000-0003-4224-8023
  
date: "February 10, 2020"
##bibliography: library.bib
tags: oceanography, remote-sensing, sattelite-imagery, spatial-data, sst, chla, ggplot, mapping, bathymetry


output: 
  uswebr::html_uswds:
    number_sections: FALSE
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

### Getting started:

```{r}
## install or load package
# install.packages("oceanmap")
library("oceanmap")

## Package overview:
?oceanmap 
help(package="oceanmap") ## list of functions
```

# Introduction
Visualizing data is a crucial step in analyzing and exploring data. During the last
two decades the statistical programming language R has become a major tool for data
analyses and visualization across different fields of science. However, creating figures
ready for scientific publication can be a tricky and time consuming task.
The oceanmap package provides some helpful functions to facilitate and optimize the
visualization of geographic and oceanographic data, such as satellite and bathymetric
data sets. Its major functions are

* plotmap() and ggplotmap() to plot land masks
* set.colorbar(), to define the colorbar (position, ticks, colormap)
* v(), an optimized tool for creating image or contour plots from different data formats (matrix, array, raster, netcdf, binary)
* figure(), to facilitate resizing and saving of figures in diverse formats (jpeg, png, eps, pdf and eps)

These functions were written in a way that they do not require a large amount of their numerous arguments to be specified but still return nice plots. Each of these functions
will be subsequently discussed.
A closely related problem represents the extraction of spatial data at defined positions/areas. Some examples on this will be discussed in an extra section.

# The plotmap-function
Often it is required to add a land mask as geographical reference to an image- or scatter-
plot. To add a landmask it is required to provide information about the region extent.
As for the entire oceanmap-package, the idea of the plotmap-function is to reduce the
user effort when creating figures. To allow a high level of user flexibility plotmap can
produce a may by 5 different ways. By providing the required region extent as:

1. geographical coordinates (longitude and latitude),
2. a raster-object
3. an extent-object
4. a region-keyword
5. or add a land mask to an existing plot

Many additional arguments can passed by the user. The idea of a region-keyword
(option 4) is to use a short character string instead of requiring axes limits. This keyword
feature is also implemented in the v-function and permits here the rapid visualization
of subregions of interest (e.g. from a raster object). Region keywords can be set up by
the add.region-function. The latter function is interactive and also defines the default
colorbar placement that is further used by the v-function.


